Human Factors Influencing Visual Statistical Inference

### 2013-06-15 sociology_paper.tex
last update before sending

Effect of graduate degree is estimated as 0.182 which is highly significant. For an 18-25 year old female in the United States, the expected difference in the probability of correct response, if she has a graduate degree as compared to a high school degree, is, for a lineup with random effect of zero:
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.042.$$  
For a lineup with a predicted random effect of one standard deviation, i.e. we move away one standard deviation of lineup variation by adding 2.293 to each of the exponent terms in the equation, this difference reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. 
A similar calculation shows the same negligible impact of age level 36-40 (0.0533 at most)  and country (0.0424 at most) on the probability of correct response. Thus even though some of the factor levels are significant, their practical impact is negligible because of high variability of lineup difficulties.

@@

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately.  As an alternative to Model \eqref{eqn:trend_response} we examined a reduced model with attempt as a continuous covariate. But that reduced model appears to be significantly different with a $p$-value of XXX. Thus we consider the bigger Model \eqref{eqn:trend_response} with attempt as a factor variable with 10 levels. Attempt 1 is considered to be base level while fitting the model.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=6.5in]{learning_trend_subject.pdf} 
   \caption{ Mean residuals of proportion correct obtained from the model \eqref{eqn:trend_response} fitted without attempt are plotted against attempt. Linear regression lines fitted through the points are shown. None of the slopes in these three experiments are statistically significant. Subject specific variations in the slopes are observed where some of the observers did worse and some did better over the successive attempts.}
   \label{fig:learning_trend_response}
\end{figure}

@@

To visualize further how the performance in correct response improves over successive attempt, we fitted Model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. The averages of these residuals for each of the attempts are plotted against attempt as shown in Figure \ref{fig:learning_trend_response}. Least square linear regression lines are drown through the points. Even though positive slopes over the attempts are observed, none of the trends is statistically significant. Subject specific least square lines shows large overall variability. Some subjects show improvement over time and some show the decrease in performance. The subject specific slopes are very different as well.

Model \eqref{eqn:trend_time} was compared to couple of other alternatives. A bigger model with attempt as a factor was considered but they did not appear to be significantly different with $p$ value XXX. A reduced model with attempt as a continuos covariate was also considered but they were significantly different with $p$-value XXX. Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. This suggest that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

@@

The lineup variance of experiment 7 is estimated as $\hat{\sigma}^2 = 10.98$ from Model \eqref{eqn:trend_response}. But from Model \eqref{eqn:trend_time} it is estimated to be much smaller in all these experiments. This suggest that the harder lineups not necessarily took more time than easier lineups. The observers spent enough times to evaluate a easy lineup and for difficult lineup they might just give up at some point and gave the feedback.

To visualize how the time taken reduces over the successive attempts, we fitted Model \eqref{eqn:trend_time} excluding the covariate attempt from the model and computed the residuals. The averages of these residuals for each attempt are plotted in Figure \ref{fig:learning_trend_time} for all three experiments. Least square regression lines are fitted to the points excluding the first attempt. The downward trends are evident in the plots. All the slopes are highly significant. For attempt 1, we observed large positive residuals for each of the experiments. These results indicate that after working with the multiple tasks each observer became efficient which helped them finish the task faster in later attempts.

@@

The proportion of correct responses for each data plot location is shown in Figure \ref{fig:location_effect}. We observe some variability of performance for different sets of null plots even though same data plot was used for all these null sets. This may happen when for some set of null plots, one null plot appears to be more similar to the actual plot. In another set of null plots this may not happen making some lineups easier than others even though the actual data plot is the same. This pattern is evident in the figure as we see proportion correct for null plot 5 is consistently above the null plot 1 for each of the locations. The overall average proportions correct responses for each location are shown using dashed lines. As we see in the plot  the overall proportion is not varying much for different locations.

The sizes of the dots in  Figure \ref{fig:location_effect} represents the number of responses obtained for that location and null set. For some locations we have as many as 10 responses. For location 1, we did not have any response for null set 1 in one of the interaction lineups. The larger variability observed for Interaction lineups could be controlled at some extent by collecting more data for each location. Since the variability is larger when we have small number of responses as we see in the figure. For example, we have only one response for null plot 3 and 5 for Interaction effect which should lead to the extreme proportion of correct since one response would either be correct or wrong.

@@

We fitted Model \eqref{manova} to test if the mean performance vectors are similar for different locations. For this we use {\it anova()} function of {\it stats} package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 and we failed to reject the hypothesis that there is no difference in location.

@@

Observer demographics seem to have significant influence on performance in time taken to evaluate a lineup. Some variations among the factors are observed in terms of probability of correct evaluation. Age group 36-40, Countries other than India and United states, People who have a graduate degree appear to have significantly more probability to evaluate a lineup correctly. Gender does not have any effect on performance.

Individual learning trend is observed in both time taken and observer performance. Some individuals improved the performances while others showed decrease in their performances over attempts. But the overall performance of human observers does not increase through successive attempts while evaluating multiple lineups. The skill in evaluating the lineup in shorter time gets improved over successive evaluations. The earlier evaluations take significantly longer than the later evaluations. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. This investigation suggest that the skilled person may only do it faster.

@@

The subjects participating this study may not necessarily know about statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference may be higher with knowledgeable observer. Thus the variations observed with individual performances could be attributed to individual abilities in terms of their training on statistical graphics. More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of knowledge about statistical graphics.

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer. The learning trend in terms of time taken shows downward trends. At some points it should level of. It would be useful to know for how many lineups time taken levels off, since that would indicate the limit beyond which no learning occurs.

Also, a lineup with fewer or larger than 20 plots may yield different results. If the size of the lineup is much higher than 20, it is intuitive that there may be location effect of the data plot in the lineup. It is because, the observer may get tiered of scanning and make decision based on the partial scanning of the lineup. On the other hand fewer than 20 plots would give observer compare the actual plot with fewer null plots give more chance of picking actual plot as an error not as an actual plot. These are the some of the issues that needs to investigated.

@@

>>>>

Effect of graduate degree is estimated as 0.182 which is highly significant. For an 18-25 year old female in the United States, the expected difference in the probability of correct response, if she has a graduate degree as compared to a high school degree, is, for a lineup with random effect of zero:
$$ \frac {\exp(-0.683+0.182)}{1+\exp(-0.683+0.182)}-  \frac {\exp(-0.683)}{1+\exp(-0.683)} = 0.377- 0.336 =0.041.$$ 
For a lineup with a predicted random effect of one standard deviation, i.e. we move away one standard deviation of lineup variation by adding 2.293 to each of the exponent terms in the equation, this difference reduces to 0.024. For two standard deviations it further reduces to 0.003 which is practically very small. 
A similar calculation shows the same negligible impact of age level 36-40 (0.0533 at most)  and country (0.0424 at most) on the probability of correct response. Thus even though some of the factor levels are significant, their practical impact is negligible because of high variability of lineup difficulties.

@@

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiment 5, 6, and 7 separately.  As an alternative to Model \eqref{eqn:trend_response} we examined a reduced model with attempt as a continuous covariate. They did not appear to be significantly different with a $p$-value of 0.856. But we consider the bigger Model \eqref{eqn:trend_response} with attempt as a factor variable with 10 levels. Attempt 1 is considered to be base level while fitting the model.

@@

\begin{figure}[htbp] 
   \centering
%   \includegraphics[width=6.5in]{learning_trend.pdf} 
    \includegraphics[width=6.5in]{learning_trend_subject.pdf} 
   \caption{ Mean residuals of proportion correct obtained from the model \eqref{eqn:trend_response} fitted without attempt are plotted against attempt. Linear regression lines fitted through the points are shown. None of the slopes in these three experiments are statistically significant. Subject specific least square lines show large subject specific variability. Some of the observers did worse and some did better over the successive attempts.}
   \label{fig:learning_trend_response}
\end{figure}

@@

To visualize further how the performance in correct response improves over successive attempt, we fitted Model \eqref{eqn:trend_response} excluding the covariates related to attempt from the model and computed the residuals. The averages of these residuals for each of the attempts are plotted against attempt as shown in Figure \ref{fig:learning_trend_response}. Least square linear regression lines are fitted through the points. Even though positive slopes over the attempts are observed, none of the trends is statistically significant. Subject specific least square lines show large overall variability. Some subjects show improvement over time and some show the decrease in performance. The subject specific slopes are very different as well.

Model \eqref{eqn:trend_time} was compared to couple of other alternatives. A bigger model with attempt as a factor was considered but they did not appear to be significantly different with $p$ value 0.236. A reduced model with attempt as a continuos covariate was also considered but they were significantly different with $p$-value $<0.001$. Table \ref{tbl:model_result_time} presents the results of Model \eqref{eqn:trend_time}. The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that later attempt took less time for an evaluation. It is interesting to note that even though observers did not improve the performance over attempt, they became efficient in responding faster. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more times than other attempts. It is because for initial attempt the observer might have gone through instructions and became familiar with the experimental environment. Each page in the web site contains information about choice reason and the observers's confidence level. Also, the first page asks for observer Identification to be typed. The later pages of the web site was similar just the lineup was changed. Thus in the later attempts an observer does not need to spend any time for reading instructions. The model reflects that fact.

@@

The lineup variance of experiment 7 is estimated as $\hat{\sigma}^2 = 10.98$ from Model \eqref{eqn:trend_response}. But from Model \eqref{eqn:trend_time} it is estimated to be much smaller in all these experiments. This suggest that the harder lineups not necessarily took more time than easier lineups. The observers spent enough times to evaluate a easy lineup and for difficult lineup they might just give up at some point and provided the feedback.

To visualize how the time taken reduces over the successive attempts, we fitted Model \eqref{eqn:trend_time} excluding the covariate attempt from the model and computed the residuals. The averages of these residuals for each attempt are plotted in Figure \ref{fig:learning_trend_time} for all three experiments. Least square regression lines are fitted to the points excluding the first attempt. The downward trends are evident in the plots. All the slopes are highly significant. For attempt 1, we observed large positive residuals for each of the experiments. These results indicate that after working with the multiple tasks each observer became efficient which helped them finish the task faster in later attempts. Subject specific least square lines show variations in slopes and overall time taken. Even though overall skill improved, there are some subjects who took more times in later attempts.

@@

The proportion of correct responses for each data plot location is shown in Figure \ref{fig:location_effect}. We observe some variability of performance for different sets of null plots even though same data plot was used for all these null sets. This may happen when for some set of null plots, one null plot appears to be more similar to the actual plot. In another set of null plots this may not happen making some lineups easier than others even though the actual data plot is the same. This pattern is evident in the figure as we see proportion correct for null plot 5 is consistently above the null plot 1 for each of the locations. The overall average proportion of correct responses for each location are shown using dashed lines. As we see in the plot  the overall proportion is not varying much for different locations.

The sizes of the dots in  Figure \ref{fig:location_effect} represents the number of responses obtained for that location and null set. For some locations we have as many as 10 responses. For location 1, we did not have any response for null set 1 in one of the interaction lineups. We observed larger variability for interaction lineups as compared to Genotype lineups. It is because some of the interaction lineups were evaluated only once.  For example, we have only one response for null plot 3 and 5 for Interaction effect which should lead to the extreme proportion of correct since one response would either be correct or wrong. The larger variability observed for Interaction lineups can be controlled at some extent by evaluating them multiple times.

@@

We fitted Model \eqref{manova} to test if the mean performance vectors are similar for different locations. For this we use {\it anova()} function of {\it stats} package of \cite{R:2012}. The results are shown in Table \ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05. This suggest that there may be no difference in location.

@@

Observer demographics seem to have significant influence on performance in time taken to evaluate a lineup. Some variations among the factors are observed in terms of probability of correct evaluation. Age group 36-40, Countries other than India and United states, People who have a graduate degree appear to be significantly different. But their practical impact on probability to correctly evaluate a lineup is very small and in some cases very negligible as the lineup difficulties increase or decreases. Gender does not have any effect on performance.

Individual learning trend is observed in both time taken and observer performance. Some individuals improved the performances while others showed decrease in their performances over attempts. But the overall performance of the observers does not increase through successive attempts while evaluating multiple lineups. This result is very important as it suggests that the power estimated for visual inference using human subject experiment is robust and may not change if those participants are allowed to give feedback again. The skill in evaluating the lineup in shorter time gets improved over successive evaluations. The earlier evaluations take significantly longer than the later evaluations.  This investigation suggest that the skilled person may only do it faster.

@@

The subjects participating this study may not necessarily know about statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference may be higher with knowledgeable observer. The fact that a person with graduate degree performs better may be associated with this. But a graduate degree does not necessarily mean to have training on statistical graphics. More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of knowledge about statistical graphics.

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest toward the task of evaluating lineups. This assumption is made based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer. The learning trend in terms of time taken shows downward trends. At some points it should level of. It would be useful to know for how many lineups time taken levels off.

A lineup with fewer or larger than 20 plots may yield different results. If the size of the lineup is much higher than 20, it is intuitive that there may be location effect of the data plot in the lineup. It is because, the observer may get tiered of scanning and make decision based on the partial scanning of the lineup. On the other hand fewer than 20 plots will allow observer to compare the actual plot with fewer null plots giving more chance of picking actual plot as an error not as an actual plot. These are the some of the issues that require further investigation.

@@

\section{Some Plots of Exploratory Data Analysis}
